Modern Statistical Methods --- Example Sheet 2

### 2020-11-14 modern_statistical_methods/C2.tex
it's weekend, party time

We have the following bound on the prediction error:
\begin{theorem}[Slow rate] \label{thm:lasso_slow_rate}
	Assume $X$ has centred and scaled columns, and assume that $Y$ has been centred, so $Y = X\beta^0 + \eps - \bar\eps \vec 1$. Let $A > 0$ and suppose
	\[
	\lambda = A\sigma\sqrt{\frac{\log(p)}{n}}.
	\]
	Let $\hat\beta = \hat\beta_\lambda^\Rm{L}$, then with probability at least $1 - 2p^{-(A^2/2 - 1)}$ we have that the MSPE satisfies
	\[
	\frac1n \norm{X(\beta^0 - \hat\beta)}_2^2 \leq 4\lambda \norm{\beta^0}_1 = 4A\sigma\sqrt{\frac{\log(p)}{n}} \norm{\beta^0}_1. 
	\]
\end{theorem}

@@

\subsection{Concentration inequalities}
Let $W$ be any random variable and $\phi \colon \RR\to[0,\infty)$ strictly increasing. Then by Markov's inequality we have
\[
\PP(W \geq t) = \PP(\phi(W) \geq \phi(t)) \leq \frac{\EE[\phi(W)]}{\phi(t)}. 
\]
Plugging in $\phi(x) = e^{\alpha x}$ (for some $\alpha > 0$), we get
\[
\PP(W \geq t) \leq e^{-\alpha t} \EE[e^{\alpha W}] = e^{-\alpha t} M_W(\alpha).
\]
Now we can take the infimum over all $\alpha$ on the right-hand side, and we get what is called the \emph{Chernoff bound}:
\[
\PP(W \geq t) \leq \inf_{\alpha > 0} e^{-\alpha t} M_W(\alpha). 
\]

@@

@@

\subsection{Variable selection}
We consider the noiseless model $Y = X\beta^0$, where no randomness is involved. Let $S = \qty{k \mid \beta_k \neq 0}$, $N = \qty{1, \dotsc, p} \setminus S$, and assume that $S = \qty{1, \dotsc, s}$ (so $\beta_1, \dotsc, \beta_s$ are nonzero and $\beta_{s+1}, \dotsc, \beta_p$ are zero).

@@

>>>>

\subsubsection{Slow prediction error rate}
We have the following bound on the prediction error:
\begin{theorem}[Slow rate] \label{thm:lasso_slow_rate}
	Assume $X$ has centred and scaled columns, and assume that $Y$ has been centred, so $Y = X\beta^0 + \eps - \bar\eps \vec 1$. Let $A > 0$ and suppose
	\[
	\lambda = A\sigma\sqrt{\frac{\log(p)}{n}}.
	\]
	Let $\hat\beta = \hat\beta_\lambda^\Rm{L}$, then with probability at least $1 - 2p^{-(A^2/2 - 1)}$ we have that the MSPE satisfies
	\[
	\frac1n \norm{X(\beta^0 - \hat\beta)}_2^2 \leq 4\lambda \norm{\beta^0}_1 = 4A\sigma\sqrt{\frac{\log(p)}{n}} \norm{\beta^0}_1. 
	\]
\end{theorem}

@@

\subsubsection{Concentration inequalities}
Let $W$ be any random variable and $\phi \colon \RR\to[0,\infty)$ strictly increasing. Then by Markov's inequality we have
\[
\PP(W \geq t) = \PP(\phi(W) \geq \phi(t)) \leq \frac{\EE[\phi(W)]}{\phi(t)}. 
\]
Plugging in $\phi(x) = e^{\alpha x}$ (for some $\alpha > 0$), we get
\[
\PP(W \geq t) \leq e^{-\alpha t} \EE[e^{\alpha W}] = e^{-\alpha t} M_W(\alpha).
\]
Now we can take the infimum over all $\alpha$ on the right-hand side, and we get what is called the \emph{Chernoff bound}:
\[
\PP(W \geq t) \leq \inf_{\alpha > 0} e^{-\alpha t} M_W(\alpha). 
\]

@@

\subsubsection{Lasso solutions}

@@

\subsubsection{Variable selection}
We consider the noiseless model $Y = X\beta^0$, where no randomness is involved. Let $S = \qty{k \mid \beta_k \neq 0}$, $N = \qty{1, \dotsc, p} \setminus S$, and assume that $S = \qty{1, \dotsc, s}$ (so $\beta_1, \dotsc, \beta_s$ are nonzero and $\beta_{s+1}, \dotsc, \beta_p$ are zero).

@@

\subsubsection*{The compatibility condition and random design}
We will show that if $X$ is a random matrix, then the compatibility condition holds with high probability.

Define, for $\Sigma \in \RR^{p \times p}$ and $S \subseteq \qty{1, \dotsc, p}$, the quantity
\[
\phi_\Sigma^2(S) = \inf_{\beta_S \neq 0, \norm{\beta_N}_1 \leq 3 \norm{\beta_S}_1} \frac{\beta\T \Sigma\beta}{\frac1{\abs{S}}\norm{\beta_S}_1^2},
\]
so that the compatiblity constant satisfies $\phi^2 = \phi_{\hat\Sigma}^2(S)$ with $\hat\Sigma \ceq X\T X /n $ and $S$ the support of $\beta^0$.

To prove our main result, we first need a lemma:
\begin{lemma}
	Suppose $\phi_{\check\Sigma}^2(S) > 0$ and $\max_{jk} \abs{\hat\Sigma_{jk} - \check\Sigma_{jk}} \leq \phi_{\check\Sigma}^2(S) / (32\abs{S})$. Then $\phi_{\hat\Sigma}^2(S) \geq \phi_{\check\Sigma}^2(S)$. 
\end{lemma}

\begin{proof}
	Let $s = \abs{S}$ and $t \ceq \phi_{\check\Sigma}^2/(32s)$. Then by applying H\"older's inequality we obtain
	\[
	\abs{\beta\T (\check\Sigma - \hat\Sigma)\beta} \leq \norm{\beta}_1 \norm{(\check\Sigma - \hat\Sigma)\beta}_\infty \leq \norm{\beta}_1^2 \max_{j,k} \abs{\hat\Sigma_{jk} - \check\Sigma_{jk}}  \leq \norm{\beta}_1^2 t. 
	\]
	Furthermore, if $\norm{\beta_N}_1 \leq 3 \norm{\beta_S}_1$, then
	\[
	\norm{\beta}_1 = \norm{\beta_N}_1 + \norm{\beta_S}_1 \leq 4 \norm{\beta_S}_1 \leq \frac{4\sqrt{\beta\T \check\Sigma\beta}}{\phi_{\check\Sigma} / s} \implies \norm{\beta_1}^2 \leq \frac{16 \beta\T \check\Sigma\beta}{\phi_{\check\Sigma}^2/s^2},
	\]
	and therefore
	\[
	\frac12 \beta\T \check\Sigma\beta = \beta\T \check\Sigma\beta - \frac{\phi_{\check\Sigma}^2}{32s} \cdot \frac{16 \beta\T \check\Sigma\beta}{\phi_{\check\Sigma}^2/s^2} \leq \beta\T \check\Sigma \beta -  t \norm{\beta}_1^2 \leq \beta\T\check\Sigma\beta - \beta\T(\check\Sigma - \hat\Sigma)\beta = \beta\T \hat\Sigma\beta. 
	\]
	Taking the infimum over all $\beta$ with $\norm{\beta_N}_1 \leq 3 \norm{\beta_S}_1$ and $\beta_S \neq 0$, we obtain the required result. 
\end{proof}

Now we can state the main theorem. We consider a sequence of models with sample size $n$ (where $n\to\infty$), and with $p$ and $s$ increasing to $\infty$ as functions of $n$.

\begin{theorem} \label{thm:lasso_compatibility_condition_asymptotics}
	Suppose the rows of $X$ are i.i.d.\ and each entry is a mean-zero sub-Gaussian random variable with parameter $v$. Let $\Sigma^0 \ceq\EE[X\T X/n]$, and suppose that $s \sqrt{\log(p)/n} \to 0$ as $n\to\infty$. Define
	\[
	\phi^2_{\hat\Sigma, s} \ceq \min_{\abs{S} = s} \phi_{\hat\Sigma}^2(S), \quad \phi_{\Sigma^0, s}^2 \ceq \min_{\abs{S} = s} \phi_{\Sigma^0}^2(S). 
	\]
	If $\phi_{\Sigma^0, s}^2 > c > 0$ (for all $n$), then $\PP(\phi_{\hat\Sigma, s}^2 \geq \phi_{\Sigma^0, s}^2/2) \to 1$ as $n\to\infty$. 
\end{theorem}

\begin{proof}
	By the previous lemma, it suffices to show that 
	\[
	(*) = \PP(\max_{jk}\abs{\hat\Sigma_{jk} - \Sigma^0_{jk}} \geq \phi_{\Sigma^0, s}^2 / (32s)) \to 0 \quad\text{ as } n \to\infty. 
	\]
	Letting $t \ceq \phi_{\Sigma^0, s}^2/(32s)$, then 
	by a union bound and Bernstein's theorem, we have that 
	\begin{align*}
	(*) &\leq p^2 \max_{j, k} \PP\qty(\abs{\sum_{i=1}^n X_{ij} X_{ik}/n - \Sigma^0_{jk}} \geq t) \\
	&\overset\star\leq p^2 2 \exp\qty(-\frac{nt^2}{2(64 v^4 + 4v^2 t)}) \\
	&\leq c_1 \exp(-c_2 \frac{n}{s^2} + c_3 \log(p)) \to 0 \quad\text{if $s \sqrt{\log(p)/n} \to 0$ as $n\to\infty$}. 
	\end{align*}
where $\star$ follows from the fact that $X_{ij}X_{ik}$ satisfies Bernstein's condition with parameters $(8v^2, 4v^2)$. 
\end{proof}

What does the previous theorem tell us? We have a weak growth condition: for example, the theorem holds if $p = n^d$ and $s = (\sqrt n)^{1 -\alpha}$ for any $d, \alpha > 0$.

\begin{example}
	Suppose the rows of $X$ are taken form a $N_p(0, \Sigma^0)$ distribution, where $\Sigma^0$ has bounded diagonal entries and the smallest eigenvalue of $\Sigma^0$ is $c_\Rm{min} > c > 0$ (for all $n$). Then $\PP(\phi_{\hat\Sigma, s}^2 \geq \frac{c_\Rm{min}}{2}) \to 1$ as $n\to\infty$ if the growth condition is satisfied. This holds for example if $\Sigma^0 = I_p$. 
\end{example}

\begin{example}
	Suppose $X$ has i.i.d.\ entries $\PP(X_{ij} = 1) = \frac12 = \PP(X_{ij} = -1)$. Then by Hoeffding's lemma, $X_{ij}$ is sub-Gaussian, and therefore the theorem holds under appropriate conditions. 
\end{example}

\begin{example}
	Let $W$ be a matrix whose rows are vectors in a basis, and let $X$ have rows which are random subsets of $W$. This does not satisfy the conditions of \cref{thm:lasso_compatibility_condition_asymptotics}, yet the lasso still works very well in this case. 
\end{example}

Generally, the Lasso demonstrates a very low estimation if $n/p$ is much larger than $s/p$, but a very high estimation error if $s/p$ is much larger than $n/p$. Usually, there is a sharp ``phase transition'' here.

\subsubsection{Computation}
The \emph{coordinate descent} algorithm is an algorithm for minimising functions of the form $f \colon \RR^d \to \RR$, and works particularly well for functions of the form
\[
f(x) = g(x) + \sum_{j=1}^d h_j(x), 
\]
where $g$ is convex and differentiable and the $h_j \colon \RR \to \RR$ are convex and continuous. Note that the Lasso objective function is clearly of this form. 
We start with an initial guess $x^{(0)}$ for a minimiser, and then for $m = 1, 2, \dotsc$ we choose, for $i = 1, \dotsc, d$, 
\[
x^{(m)}_i = \argmin_{x_i \in \RR} f\qty(x_1^{(m)}, \dotsc, x_{i-1}^{(m)}, x_i, x_{i+1}^{(m-1)}, \dotsc, x_d^{(m-1)}). 
\]
It can be shown that for the Lasso objective function, each step of the coordinate descent algorithm admits a closed-form solution.

We have the following:
\begin{theorem}[Tseng] \label{thm:tseng} If $A_0 \ceq \qty{x \mid f(x) \leq f(x^0)}$ is compact, then every converging subsequence of $(x^{(m)})$ converges to a minimiser of $f$. 
\end{theorem}

Using this theorem, we can see the following:
\begin{corollary}
	Suppose $A^0$ is compact. Then 
	\begin{enumerate}
		\item There exists a mimiser $x^*$ of $f$, and $f(x^{(m)}) \to f(x^*)$;
		\item If $x^*$ is the unique minimiser of $f$, then $x^{(m)} \to x^*$. 
	\end{enumerate}
\end{corollary}

\begin{proof}
	Since $f$ is continuous, it attains a minimum $f(x^*)$ on $A_0$. Suppose $f(x^{(m)}) \not\to f(x^*)$. Note that $f(x^{(m)})$ is a monotone decreasing sequence, so it must converge. Furthermore, $(x^{(m)}) \subseteq A_0$, so there exists a subsequence of $(f(x^{(m)}))$ converging to $f(x^*)$ by \cref{thm:tseng}. This proves that $f(x^{(m)}) \to f(x^*)$.

	Suppose that $x^*$ is the unique minimiser of $f$, then by \cref{thm:tseng} we know that every convergent subsequence of $x^{(m)}$ converges to $x^*$. Since $x^{(m)}$ is bounded, this proves that $x^{(m)} \to x^*$ as well.  
\end{proof}

Often, we want to solve the Lasso on a grid of values $\lambda_0 > \dotsb > \lambda_L$. To speed up the coordinate descent procedure, we can use the minimiser for $\lambda_{\ell -1}$ as a starting point for coordinate descent on $\lambda_\ell$. An ``active set'' strategy can speed up computation even further: 
\begin{enumerate}
	\item Initialise $A_\ell \ceq \qty{k \mid \beta^L_{\ell - 1, k} \neq 0}$;
	\item Perform coordinate descent with $\lambda_\ell$ only on the coordinates in $A_\ell$, call this result $\hat\beta$. 
	\item Let $V$ be the set of coordinates which violate the KKT conditions;
	\item If $V = \emptyset$, set $\hat\beta_{\lambda_\ell}^L = \hat\beta$. Else, update $A_\ell = A_\ell \cup V$ and return to step 2. 
\end{enumerate}

\begin{remark}
	There are multiple possible algorithms for computing Lasso solutions, coordinate descent is simply one option. 
\end{remark}

\subsection{Extensions of the Lasso}
One of the ways to extend the Lasso is to change the loss function (for example,
the logistic loss function can be used if $Y_i \in \qty{-1, 1}$). 
There are other extensions.

\subsubsection{Structural penalties}
\begin{example}[Group lasso]
	Let $G_1, \dotsc, G_q$ be a partition of $\qty{1, \dotsc, p}$, then instead of the Lasso penalty we can use
	\[
	\lambda \sum_{j=1}^q m_j \norm{\beta_{G_j}}_2,
	\]
	where $m_j$ is typically chosen as $\sqrt{\abs{G_j}}$. This estimator will generally ensure that either the entire group $G_j$ is shrunk to 0 or not at all. 
\end{example}

\begin{example}[Fused lasso]
	Suppose the $(\beta_j)$ are ordered, for example if $j$ fixes time. If we expect $(\beta_j)$ to be piecewise constant (or at least that $\beta^0_j$ is close to $\beta^0_{j+1}$), we can use the penalty 
	\[
	\lambda_1 \sum_{j=1}^{p-1} \abs{\beta_j - \beta_{j+1}} + \lambda_2 \norm{\beta}_1, 
	\]
	where the term $\lambda_2 \norm{\beta}_1$ can be omitted if shrinkage is not desired. 
\end{example}

\subsubsection{Debiasing the Lasso}
A useful property of the Lasso is that non-significant coefficients can be shrunk towards 0, but this same effect also causes many significant coefficients to be shrunk  towards 0. A possible solution is to only use the Lasso to estimate which coefficients are 0 and then use a different estimator (i.e., OLS) for the other coefficients.

Another option is to use a non-convex penalty function which does not shrink large values of $\hat\beta$ as much.
